Students' Alcohol Consumption Impacts on Academic Results on the Example of a Secondary Scool Maths Class in Portugal

### 2016-10-13 knit_MSproject.Rmd
Knit-able version of MSproject. + However, some code are not working.

---
title: "Students' Alcohol Consumption impacts on academic Results on the example of a secondary Scool Maths Class in Portugal"
output: github_document
Data set source: P. Cortez and A. Silva. Using Data Mining to Predict Secondary School Student Performance. In A. Brito and J. Teixeira Eds., Proceedings of 5th FUture BUsiness TEChnology Conference (FUBUTEC 2008) pp. 5-12, Porto, Portugal, April, 2008, EUROSIS, ISBN 978-9077381-39-7. 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Students' Alcohol Consumption impacts on academic Results on the example of a secondary Scool Maths Class in Portugal

---
Content: 
1. Introduction

2. Dataset Description

3. Descriptive Statistics

4. Correlation Analysis

5. Hypothesis Testing

6. Regression

7. Conclusion

8. References

9. Appendices
---

## Introduction

Recent events in the world economy have lead to the decline in economic performance in many countries, amongst which is the decline of econommic growth in the European Union (EU).  As a response to these changes,  the EU has made it a priority to reverse this trend through investments that would provide new jobs and create a better educated workforce. Hence, the European Commission (EC) has set up a ‘Strategy for smart, sustainable and inclusive growth’ (European Comission, 2010), where a crucial part of investment is to contribute to the education of the next generations: The aim is to reduce the share of early school leavers to below 10% and ensure that at least 40% of the younger generation obtain a tertiary degree.

@@

## Description of Dataset and Descriptive Statistics

```{r pressure, include=FALSE}
sm <- read.csv("student-mat.csv", sep=';')

# Descriptive statistics
library(ggplot2)

@@

blank_theme <- theme_minimal()+ theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid=element_blank(),
    axis.ticks = element_blank(),
    plot.title=element_text(size=14, face="bold")
  )
########################################################
# Sex
summary(sm[,'sex']) 
#F=208 #M=187

df_gender <- data.frame(Gender=c("Female", "Male"), value=c(208, 187))
pie_gender <- ggplot(data=df_sex, aes(x="", y=value, fill=Gender)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4)

###
# Address
summary(sm[,'address']) 
#Rural = 77 #Urban = 307

df_address <- data.frame(Address=c("Rural", "Urban"), value=c(77, 307))
pie_address <- ggplot(data=df_address, aes(x="", y=value, fill=Address)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4)

###
#Study Time

studyTime <- table(sm$studytime)
#  1   2   3   4 
# 105 198  65  27

df_studyTime <- data.frame(Study_Time=c('<2 hours', "2 to 5 hours", "5 to 10 hours", ">10 hours"), value=c(105, 198, 65, 27))
#weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)

bar_studytime <- ggplot(data=df_studyTime, aes(x="", y=value, fill=Study_Time)) + 
  geom_bar(width = 1, stat = "identity") + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4)

@@

failures <- table(sm$failures)
# 0   1   2   3 
# 312  50  17  16 
df_failures <- data.frame(Failures_Times=c("0", "1", "2", "3"), value=c(312, 50, 17, 16))

pie_failures <- ggplot(data=df_failures, aes(x="", y=value, fill=Failures_Times)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4)

@@

df_school <- data.frame(Schools=c("Gabriel Pereira", "Mousinho da Silveira"), value=c(349,46))
bar_school <- ggplot(data=df_school, aes(x="", y=value, fill=Schools)) + 
  geom_bar(width = 1, stat = "identity") + 
  blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4)

###
# Father's Job
summary(sm[,'Fjob'])
ggplot(data=sm) + geom_bar(aes((sm[,'Fjob']))) + labs(x="Father's Job")
# at_home   health  other  services  teacher 
#   20       18      217      111       29

df_fjob <- data.frame(Group = c("Health", "Work at Home", "Teacher", "Servies", "Other"), value = c(18, 20, 29, 111, 217))
pie_fjob <- ggplot(data = df_fjob, aes(x="", y=value, fill=Group)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme +
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4) +
  ggtitle("Father's Job")

###
# Mother's Job
summary(sm[,'Mjob'])
ggplot(data=sm) + geom_bar(aes((sm[,'Mjob']))) + labs(x="Mother's Job")
# at_home   health  other  services  teacher 
# 59       34        141      103       58

df_mjob <- data.frame(Group = c("Health", "Teacher", "Work at Home", "Services", "Other"), value = c(34, 58, 59, 103, 141))
pie_mjob <- ggplot(data = df_mjob, aes(x="", y=value, fill=Group)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme +
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4) +
  ggtitle("Mother's Job")

###
# Student's Guardian
summary(sm[,'guardian'])
ggplot(data=sm) + geom_bar(aes(guardian)) + labs(x="Student's Guardian")
# father mother  other 
# 90    273     32

df_guardian <- data.frame(group = c("Father", "Mother", "Other"), value = c(90, 273, 32))
pie_guardian <- ggplot(data = df_guardian, aes(x="", y=value, fill=group)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme +
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4) +
  ggtitle("Guardian")

###############################################################################################

@@

#### 1. General Information
General information about the students include gender, address and age. From the descriptive statistics (section 3) we can derive that the age from the students in this dataset ranges from 15 to 22 years with an average of 16.7 years. Out of the total of 395, there are 52.7% female and 47.3% male students. Most students would live in an urban environment, while approx. 20% live in an rural environment.

```{r, echo=FALSE}
pie_gender

pie_address
```

#### 2. Family  
Information in the students families includes the family size, educational background of their parents and their jobs, and the student's guardian.

```{r, echo=FALSE}
pie_fjob

pie_mjob

pie_guardian
```

#### 3. Education
The education section of this dataset covers the split between the two schools observed, why each student has chosen that particular school, and how long it takes each student to get to school. In addition, it provides information on the student's study time, extra educational support or classes taken, whether they have failed to pass any classes and if the student is involved in extra curricular activities. Finally, there is data on their previous education and aspirations on whether to go into higher education.

```{r, echo=FALSE}
bar_school
bar_studytime

@@

#### 4. Social Life and Health 
The social life section provieds information on activities the student is involved in outside of school. This includes

#### 5. Grades
The grades are split into three grades: the first period (G1), second period (G2) and final grade (G3). For the purpose of thy analysis, G1 and G2 will be neglected. 
```{r, echo=FALSE}
pie_failures

## Correlation Analysis

Text

```{r pressure, echo=FALSE}
plot(pressure)
```
```{r}
sm <- student.mat
for (i in 16:24) {                                  
  sm[,i] <- as.character(sm[,i])
  for (j in 1:nrow(sm)) {
    if (sm[j,i] == "yes") sm[j,i] <- 1
    if (sm[j,i] == "no") sm[j,i] <- 0
  }
  sm[,i] <- as.numeric(sm[,i])
}

#General test 
```{r}
t.test(sm[,"Walc"], sm[,"Dalc"])
library(ggplot2)
ggplot(data = sm) + geom_bar(aes(x=age))
ggplot(data = sm) + geom_bar(aes(x=sex))
ggplot(data = sm) + geom_bar(aes(x=Dalc))
ggplot(data = sm) + geom_bar(aes(x=Walc))
i <- ggplot(sm, aes(x = sm[,"age"], y = sm[,"sex"]))
i + geom_bin2d(binwidth = c(1,1)) #LABEL #Distribution of gender and age
i <- ggplot(sm, aes(x = sm[,"Dalc"], y = sm[,"Walc"]))
i + geom_bin2d(binwidth = c(1,1)) #LABEL #Distribution of Walc and Dalc

#Test and plot on gender and age

###age
```{r}
f <- ggplot(data =sm, aes(x = sm[,"Dalc"], y = sm[,"age"]))
f + geom_smooth(method = lm) #LABEL
f <- ggplot(data =sm, aes(x = sm[,"Walc"], y = sm[,"age"]))
f + geom_density(method = lm)  #LABEL
f + geom_smooth(na.rm = TRUE)
i <- ggplot(sm, aes(x = sm[,"age"], y = sm[,"sex"]))
i + geom_bin2d(binwidth = c(1,1))
i <- ggplot(sm, aes(x = sm[,"absences"], y = sm[,"Walc"]))
i + geom_bin2d(binwidth = c(1,1))
cor.test(sm[,"age"],sm[,"Dalc"])
cor.test(sm[,"age"],sm[,"Walc"])
```

###gender
```{r}
for (i in 2) {                                  
  sm[,i] <- as.character(sm[,i])
  for (j in 1:nrow(sm)) {
    if (sm[j,i] == "M") sm[j,i] <- 1
    if (sm[j,i] == "F") sm[j,i] <- 0
  }
  sm[,i] <- as.numeric(sm[,i])
}
cor.test(sm[,"sex"],sm[,"Walc"])
cor.test(sm[,"sex"],sm[,"Dalc"])
```

##Correlations tests 
As inputs have to be numeric it requires less lines to split the code that way instead of doing block per parameters analysed (Failures, paid, Medu, ... )
We could have done for loops here to calculate several correlations but it the results wasn't user friendly and doint it that way gave some useless correlations.
```{r}
sm <- as.numeric(sm)
sm1 <- subset(sm, sex == 1) #Create subsets with boys or girls
sm1 <- subset(sm, sex == 0)
```

###Romantic
```{r}
cor.test(sm[,"romantic"], sm[,"Dalc"])
cor.test(sm[,"romantic"], sm[,"Walc"])
```

###absenses
```{r}
cor.test(sm[,"absences"], sm[,"Dalc"])
cor.test(sm[,"absences"], sm[,"Walc"])
```

###Failures
```{r}
cor.test(sm[,"failures"], sm[,"Dalc"])
cor.test(sm[,"failures"], sm[,"Walc"])
```

###Boypart
```{r}
cor.test(sm1[,"failures"], sm1[,"Dalc"])
cor.test(sm1[,"failures"], sm1[,"Walc"])
```

###Medu
```{r}
cor.test(sm[,"Medu"], sm[,"Dalc"])
cor.test(sm[,"Medu"], sm[,"Walc"])
```

###Fedu
```{r}
cor.test(sm[,"Fedu"], sm[,"Dalc"])
cor.test(sm[,"Fedu"], sm[,"Walc"])
```

###Travel Time
```{r}
cor.test(sm[,"traveltime"], sm[,"Dalc"])
cor.test(sm[,"traveltime"], sm[,"Walc"])
```

###famrel
```{r}
cor.test(sm[,"famrel"], sm[,"Dalc"])
cor.test(sm[,"famrel"], sm[,"Walc"])
```

###goout
```{r}
cor.test(sm[,"goout"], sm[,"Dalc"])
cor.test(sm[,"goout"], sm[,"Walc"])
```

###famsize
```{r}
cor.test(sm[,"famsize"], sm[,"Dalc"])
cor.test(sm[,"famsize"], sm[,"Walc"])
```

###internet
```{r}
cor.test(sm[,"internet"], sm[,"Dalc"])
cor.test(sm[,"internet"], sm[,"Walc"])
```

###famrel
```{r}
cor.test(sm[,"famrel"], sm[,"Dalc"])
cor.test(sm[,"famrel"], sm[,"Walc"])
```

###Grades
```{r}
cor.test(sm[,"G1"], sm[,"Dalc"])
cor.test(sm[,"G1"], sm[,"Walc"])
cor.test(sm[,"G2"], sm[,"Dalc"])
cor.test(sm[,"G2"], sm[,"Walc"])
cor.test(sm[,"G3"], sm[,"Dalc"])
cor.test(sm[,"G3"], sm[,"Walc"])
cor.test(sm[,"G3"], sm[,"goout"])
cor.test(sm[,"G3"], sm[,"failures"])
cor.test(sm[,"G3"], sm[,"absences"])
```

#EARLY SIGNS OF ACADEMIC FAILURES
###abscences
```{r}
cor.test(sm[,"absences"], sm[,"Dalc"])
cor.test(sm[,"absences"], sm[,"Walc"])
cor.test(sm1[,"absences"], sm1[,"Dalc"])
cor.test(sm1[,"absences"], sm1[,"Walc"])
cor.test(sm2[,"absences"], sm2[,"Dalc"])
cor.test(sm2[,"absences"], sm2[,"Walc"])
```

###Failures
```{r}
cor.test(sm[,"failures"], sm[,"Dalc"])
cor.test(sm[,"failures"], sm[,"Walc"])
cor.test(sm1[,"failures"], sm1[,"Dalc"])
cor.test(sm1[,"failures"], sm1[,"Walc"])
cor.test(sm2[,"failures"], sm2[,"Dalc"])
cor.test(sm2[,"failures"], sm2[,"Walc"])
```

###goout
```{r}
cor.test(sm[,"goout"], sm[,"Dalc"])
cor.test(sm[,"goout"], sm[,"Walc"])
cor.test(sm1[,"goout"], sm1[,"Dalc"])
cor.test(sm1[,"goout"], sm1[,"Walc"])
cor.test(sm2[,"goout"], sm2[,"Dalc"])
cor.test(sm2[,"goout"], sm2[,"Walc"])
```

```{r}
sm <- student.mat

###remi text
```{r}
layout(matrix(1:14, ncol = 7))
cdplot(sm[,"guardian"] ~ sm[,"Dalc"], data = sm, xlab = "Alcohol consumption during working days ", ylab = "Guardian", border = 1)
cdplot(sm[,"guardian"] ~ sm[,"Walc"], data = sm, xlab = "Alcohol consumption during Week-end ", ylab = "Guardian", border = 1)
cdplot(sm[,"Medu"] ~ sm[,"Dalc"], data = sm, xlab = "Alcohol consumption during working days ", ylab = "Father's education", border = 1)
cdplot(sm[,"Medu"] ~ sm[,"Walc"], data = sm, xlab = "Alcohol consumption during Week-end ", ylab = "Father's education", border = 1)
cdplot(sm[,"Fedu"] ~ sm[,"Dalc"], data = sm, xlab = "Alcohol consumption during working days ", ylab = "Mother's education", border = 1)
cdplot(sm[,"Fedu"] ~ sm[,"Walc"], data = sm, xlab = "Alcohol consumption during working Week-end ", ylab = "Mother's education", border = 1)
cdplot(sm[,"Pstatus"] ~ sm[,"Dalc"], data = sm, xlab = "Alcohol consumption during working days ")
cdplot(sm[,"Pstatus"] ~ sm[,"Walc"], data = sm, xlab = "Alcohol consumption during Week-end ")
cdplot(sm[,"famsize"] ~ sm[,"Dalc"], data = sm, xlab = "Alcohol consumption during working days ")
cdplot(sm[,"famsize"] ~ sm[,"Walc"], data = sm, xlab = "Alcohol consumption during working Week-end ")
cdplot(sm[,"Mjob"] ~ sm[,"Dalc"], data = sm, xlab = "Alcohol consumption during working days")
cdplot(sm[,"Mjob"] ~ sm[,"Walc"], data = sm, xlab = "Alcohol consumption during working Week-end")
cdplot(sm[,"famrel"] ~ sm[,"Dalc"], data = sm, xlab = "Alcohol consumption during working days")
cdplot(sm[,"famrel"] ~ sm[,"Walc"], data = sm, xlab = "Alcohol consumption during working Week-end ")
```

###remi text
```{r}
layout(matrix(1:10, ncol = 5))
cdplot(sm[,"romantic"] ~ sm[,"Dalc"], data = sm)
cdplot(sm[,"romantic"] ~ sm[,"Walc"], data = sm)
cdplot(sm[,"goout"] ~ sm[,"Dalc"], data = sm)
cdplot(sm[,"goout"] ~ sm[,"Walc"], data = sm)
cdplot(sm[,"internet"] ~ sm[,"Dalc"], data = sm)
cdplot(sm[,"internet"] ~ sm[,"Walc"], data = sm)
cdplot(sm[,"activities"] ~ sm[,"Dalc"], data = sm)
cdplot(sm[,"activities"] ~ sm[,"Walc"], data = sm)
cdplot(sm[,"freetime"] ~ sm[,"Dalc"], data = sm)
cdplot(sm[,"freetime"] ~ sm[,"Walc"], data = sm)
```

###remi text
```{r}
layout(matrix(1:8, ncol = 4))
cdplot(sm[,"failures"] ~ sm[,"Dalc"], data = sm)
cdplot(sm[,"failures"] ~ sm[,"Walc"], data = sm)
cdplot(sm[,"schoolsup"] ~ sm[,"Dalc"], data = sm)
cdplot(sm[,"schoolsup"] ~ sm[,"Walc"], data = sm)
cdplot(sm[,"studytime"] ~ sm[,"Dalc"], data = sm)
cdplot(sm[,"studytime"] ~ sm[,"Walc"], data = sm)
cdplot(sm[,"G3"] ~ sm[,"Dalc"], data = sm)
cdplot(sm[,"G3"] ~ sm[,"Walc"], data = sm)

##Some regression
```{r}
RegFails <- lm (Walc ~ failures, data = sm)
summary (lm (Walc ~ failures, data = sm))
libraryr(ggplot2)
f <- ggplot(data =sm, aes(x = sm[,"Walc"], y = sm[,"failures"]))
f + geom_smooth(method = lm)
libraryr(ggplot2)
f <- ggplot(data =sm, aes(x = sm[,"Dalc"], y = sm[,"failures"]))
f + geom_smooth(method = lm)

#T
###GoOUT
```{r}
sm[,"goout"] <- as.numeric(sm[,"goout"])
sm[,"Dalc"] <- as.numeric(sm[,"Dalc"])
sm[,"Walc"] <- as.numeric(sm[,"Walc"])
cor.test(sm[,"goout"], sm[,"Dalc"])
cor.test(sm[,"goout"], sm[,"Walc"]) #Stong correlation
RegGoout <- lm (goout ~ Walc, data = sm)
RegGoout
libraryr(ggplot2)
f <- ggplot(data =sm, aes(x = sm[,"goout"], y = sm[,"Walc"]))
f + geom_smooth(method = lm) #TBS
sm[,"goout"] <- as.factor(sm[,"goout"])
sm[,"Dalc"] <- as.factor(sm[,"Dalc"])
sm[,"Walc"] <- as.factor(sm[,"Walc"])
layout(matrix(1:2, ncol = 2))
cdplot(sm[,"goout"] ~ sm[,"Dalc"], data = sm)
cdplot(sm[,"goout"] ~ sm[,"Walc"], data = sm)

###Internet
```{r}
sm[,"internet"] <- as.numeric(sm[,"internet"])
sm[,"Dalc"] <- as.numeric(sm[,"Dalc"])
sm[,"Walc"] <- as.numeric(sm[,"Walc"])
cor.test(sm[,"internet"], sm[,"Dalc"])
cor.test(sm[,"internet"], sm[,"Walc"])
```

###Activities
```{r}
sm[,"activities"] <- as.numeric(sm[,"activities"])
sm[,"Dalc"] <- as.numeric(sm[,"Dalc"])
sm[,"Walc"] <- as.numeric(sm[,"Walc"])
cor.test(sm[,"activities"], sm[,"Dalc"])
cor.test(sm[,"activities"], sm[,"Walc"])

###Freetime
```{r}
sm[,"freetime"] <- as.numeric(sm[,"freetime"])
sm[,"Dalc"] <- as.numeric(sm[,"Dalc"])
sm[,"Walc"] <- as.numeric(sm[,"Walc"])
cor.test(sm[,"freetime"], sm[,"Dalc"])
cor.test(sm[,"freetime"], sm[,"Walc"])
```

```{r}
RegAbs <- lm (absences ~ Walc, data = sm)
RegAbs
library(ggplot2)
f <- ggplot(data =sm, aes(x = Walc, y = absences))
f + geom_smooth(method = lm) #TBS
sm[,"absences"] <- as.factor(sm[,"absences"])
sm[,"Dalc"] <- as.factor(sm[,"Dalc"])
sm[,"Walc"] <- as.factor(sm[,"Walc"])
layout(matrix(1:2, ncol = 2))
cdplot(sm[,"absences"] ~ sm[,"Dalc"], data = sm)
cdplot(sm[,"absences"] ~ sm[,"Walc"], data = sm)
```

##Mean tests
```{r}
mean(sm[sm[, "Fedu"] == "5", "Dalc"]) - mean(sm[sm[, "Medu"] == "5", "Dalc"])
mean(sm[sm[, "guardian"] == "mother", "Dalc"]) - mean(sm[sm[, "guardian"] == "father", "Dalc"])

Text

```{r pressure, echo=FALSE}
plot(pressure)
```

## Correlation and Regression

In this report, we would like to find factors that are related to students academic performance. We will perform a multivariate regression analysis and use step-wise variable selection to find significant variables related to our response variable (academic performance - "Score"). Prior to the regression analysis, we want to get a general idea about the relationship between the variables. There are several things we want to inspect :

1) Distribution of response variable
if there is skewness in the distribution, we should consider transforming the variable (i.e logarithmic)

2) Correlation between response variable and each explanatory variables

3) Correlation among explanatory variables
if there is a high correlation among explanatory variables, then there is possible multicollinearity problem

I. Distribution of Response Variable
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
ggplot(sm) + geom_density(aes(x=Score)) + labs(x = "Academic Score", y="% Frequency", title = "Academic Score of Secondary School Student")
```
From the graph, we can see that the distribution is slightly left-skewed. Generally, we would transform the variable to become more normally distributed. However, for the purpose of this project, we will not do any transfomation on the response variable and assume that the distribution is normal.

>>>>

---
title: "Students' Alcohol Consumption Impacts on Academic Results on the Example of a Secondary Scool Maths Class in Portugal"
author: "Group 2"
date: "13 October 2016"
output: html_document
Data set source: P. Cortez and A. Silva. Using Data Mining to Predict Secondary School Student Performance. In A. Brito and J. Teixeira Eds., Proceedings of 5th FUture BUsiness TEChnology Conference (FUBUTEC 2008) pp. 5-12, Porto, Portugal, April, 2008, EUROSIS, ISBN 978-9077381-39-7.

---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

Content:

1. Introduction

2. Dataset Description and Descriptive Statistics

    + a. General Information
    + b. Family
    + c. Education 
    + d. Social Life and Health
    + e. Grades  \n

3. Hypothesis Testing

4. Correlation Analysis and Regression

5. Interesting Findings

6. Conclusion

7. References

8. Appendices

## 1. Introduction 
Recent events in the world economy have lead to the decline in economic performance in many countries, amongst which is the decline of econommic growth in the European Union (EU).  As a response to these changes,  the EU has made it a priority to reverse this trend through investments that would provide new jobs and create a better educated workforce. Hence, the European Commission (EC) has set up a ‘Strategy for smart, sustainable and inclusive growth’ (European Comission, 2010), where a crucial part of investment is to contribute to the education of the next generations: The aim is to reduce the share of early school leavers to below 10% and ensure that at least 40% of the younger generation obtain a tertiary degree.

@@

## 2. Dataset Description and Descriptive Statistics 
```{r a, include=FALSE}
# Read csv file
sm <- read.csv("/Users/Felix/Desktop/R_project/student-mat.csv", sep=';')

# Descriptive statistics
library(ggplot2)
library(gridExtra)

@@

blank_theme <- theme_minimal()+ theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    panel.border = element_blank(),
    panel.grid=element_blank(),
    axis.ticks = element_blank(),
    plot.title=element_text(size=14, face="bold")
  )
########################################################
# Sex
summary(sm[,'sex']) 
#F=208 #M=187
df_gender <- data.frame(Gender=c("Female", "Male"), value=c(208, 187))
pie_gender <- ggplot(data=df_gender, aes(x="", y=value, fill=Gender)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + 
  c(0, cumsum(value)[-length(value)]), label = (value)), size=4, show.legend=FALSE) +
  ggtitle("Student's Gender")

###
# Address
summary(sm[,'address']) 
#Rural = 77 #Urban = 307
df_address <- data.frame(Type=c("Rural", "Urban"), value=c(77, 307))
pie_address <- ggplot(data=df_address, aes(x="", y=value, fill=Type)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4) +
  ggtitle("Student's Home Address Type")

###
#Study Time
#weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)
studyTime <- table(sm$studytime)
#  1   2   3   4 
# 105 198  65  27

df_studyTime <- data.frame(Hours=c('<2 hours', "2 to 5 hours", "5 to 10 hours", ">10 hours"), value=c(105, 198, 65, 27))
bar_studytime <- ggplot(data=df_studyTime, aes(x="", y=value, fill=Hours)) + 
  geom_bar(width = 1, stat = "identity") + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4) +
  ggtitle('Study Time')

@@

failures <- table(sm$failures)
# 0   1   2   3 
# 312  50  17  16 
df_failures <- data.frame(Times=c("3", "2", "1", "0"), value=c(16, 17, 50, 312))
pie_failures <- ggplot(data=df_failures, aes(x="", y=value, fill=Times)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4) +
  ggtitle('Number of Past Class Failures')

@@

df_school <- data.frame(Group=c("Gabriel Pereira", "Mousinho da Silveira"), value=c(349,46))
bar_school <- ggplot(data=df_school, aes(x="", y=value, fill=Group)) + 
  geom_bar(width = 1, stat = "identity") + 
  scale_fill_brewer() + blank_theme + 
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=4) +
  ggtitle('School')

###
# Father's Job
summary(sm[,'Fjob'])
# at_home   health  other  services  teacher 
#   20       18      217      111       29

df_fjob <- data.frame(Job = c("Health", "Work at Home", "Teacher", "Servies", "Other"), value = c(18, 20, 29, 111, 217))
pie_fjob <- ggplot(data = df_fjob, aes(x="", y=value, fill=Job)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme +
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=3) +              ggtitle("Father's Job")

###
# Mother's Job
summary(sm[,'Mjob'])
# at_home   health  other  services  teacher 
# 59       34        141      103       58

df_mjob <- data.frame(Job = c("Health", "Teacher", "Work at Home", "Services", "Other"), value = c(34, 58, 59, 103, 141))
pie_mjob <- ggplot(data = df_mjob, aes(x="", y=value, fill=Job)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme +
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=3) +
  ggtitle("Mother's Job")

###
# Student's Guardian
summary(sm[,'guardian'])
# father mother  other 
# 90    273     32

df_guardian <- data.frame(Type = c("Father", "Mother", "Other"), value = c(90, 273, 32))
pie_guardian <- ggplot(data = df_guardian, aes(x="", y=value, fill=Type)) + 
  geom_bar(width = 1, stat = "identity") + 
  coord_polar("y", start=0) + 
  scale_fill_brewer() + blank_theme +
  theme(axis.text.x=element_blank()) +
  geom_text(aes(y = value/3 + c(0, cumsum(value)[-length(value)]), label = (value)), size=3) +
  ggtitle("Guardian")
```

#### a. General Information
General information about the students include gender, address and age. From the descriptive statistics (section 3) we can derive that the age from the students in this dataset ranges from 15 to 22 years with an average of 16.7 years. Out of the total of 395, there are 52.7% female and 47.3% male students. Most students would live in an urban environment, while approx. 20% live in an rural environment. 
```{r b, echo=FALSE}
grid.arrange(pie_gender, pie_address, ncol=2)
```
#### b. Family  
Information in the students families includes the family size, educational background of their parents and their jobs, and the student's guardian.

```{r c, echo=FALSE}
grid.arrange(pie_guardian, pie_fjob, pie_mjob, widths = c(1, 1, 2), layout_matrix = rbind(c(1,1,1), c(2,2,3)))
```

#### c. Education
The education section of this dataset covers the split between the two schools observed, why each student has chosen that particular school, and how long it takes each student to get to school. In addition, it provides information on the student's study time, extra educational support or classes taken and if the student is involved in extra curricular activities. Finally, there is data on their previous education and aspirations on whether to go into higher education.

```{r d, echo=FALSE}
grid.arrange(bar_studytime, bar_school, ncol=2)

@@

#### d. Social Life and Health 
The social life section provieds information on activities the student is involved in outside of school. This includes how much leisure time they have, whether they are in a relationship and how often they go out. Moreover, there is data on how much alcohol students consume and in what health condition they are in.

#### e. Grades
The grades are split into three grades: the first period (G1), second period (G2) and final grade (G3). For the purpose of thy analysis, G1 and G2 will be neglected. In addition, there is information on whether they have failed to pass any classes. 
```{r e, echo=FALSE}
pie_failures
```

## 3. Hypothesis Testing

## 4. Correlation Analysis and Regression
In this report, we would like to find factors that are related to students academic performance. We will perform a multivariate regression analysis and use step-wise variable selection to find significant variables related to our response variable (academic performance - "Score"). Prior to the regression analysis, we want to get a general idea about the relationship between the variables. There are several things we want to inspect:

a. Distribution of response variable

if there is skewness in the distribution, we should consider transforming the variable (i.e logarithmic)

b. Correlation between response variable and each explanatory variables

c. Correlation among explanatory variables

if there is a high correlation among explanatory variables, then there is possible multicollinearity problem

### a. Distribution of Response Variable
```{r f, echo=TRUE}
knitr::opts_chunk$set(echo = FALSE)

sm[,"Score"] <- (sm[,"G1"] + sm[,"G2"] + sm[,"G3"])/3

ggplot(sm) + geom_density(aes(x=Score)) + labs(x = "Academic Score", y="% Frequency", title = "Academic Score of Secondary School Student")
```
From the graph, we can see that the distribution is slightly left-skewed. Generally, we would transform the variable to become more normally distributed. However, for the purpose of this project, we will not do any transfomation on the response variable and assume that the distribution is normal.

### b. Correlation between response variable and each explanatory variables
```{r g, echo=TRUE}
knitr::opts_chunk$set(echo = FALSE)

panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * r)
}

#below - too large to knit

#pairs(sm[,c("G3", "G1","G2","school","sex","age", "address","famsize","Pstatus","Medu","Fedu",
#            "Mjob","Fjob","reason","guardian","traveltime","studytime","failures",
#            "schoolsup","famsup","paid","activities","nursery","higher","internet",
#            "romantic","famrel","freetime","goout","Dalc","Walc","health","absences")],
#      lower.panel = panel.smooth, upper.panel = panel.cor)

@@

### c. Correlation between response variable and each explanatory variables
```{r h, echo=TRUE}
knitr::opts_chunk$set(echo = FALSE)

panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits = digits)[1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = cex.cor * r)
}

# too large to knit

#pairs(sm[,c("G3", "G1","G2","school","sex","age", "address","famsize","Pstatus","Medu","Fedu",
#            "Mjob","Fjob","reason","guardian","traveltime","studytime","failures",
#            "schoolsup","famsup","paid","activities","nursery","higher","internet",
#            "romantic","famrel","freetime","goout","Dalc","Walc","health","absences")],
#      lower.panel = panel.smooth, upper.panel = panel.cor)
```

```{r i, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#2. derive variables
